### 2023-08-08 chapter_2/src/manuscript.qmd
update years for empirical chapter 1, study 2

This chapter applies the snap-judgment model to a simple case. The use of color on yard signs used by campaigns in elections. I focus on yard signs as they are a common form of campaign marketing and evidence suggests that they can shape individual behavior [@makse_et-al_2019_oup] and electoral outcomes [@green_et-al_2016_es]. Additionally, I start by examining the colors red and blue. While there are other colors that may provide political information -- black and white for Black Lives Matter and Blue Lives Matter movements, the overarching goal of this book is to lay the foundation for a larger research agenda in visual political psychology [see @bucy_joo_2021_ijpp]. Without such foundations, we are studying complex cases and circumstances upon assumptions without a grasp of how, and if they even do, occur in the cases that might seem trivial to some. While there indeed is empirical and experiential evidence suggesting that the Republicans use the color red and Democrats use the color blue [@schloss_palmer_2014_pbr;@williams_et-al_2022_jomp], there is yet to be a theoretical framework that explains how these connections happen at the individual level. It may also be unsatisfying that I do not address *how* these connections came about. While the connections between Republicans with the color red and Democrats with the color blue seemed to quicken after the 2000 Presidential election where news networks covering the election began to consistently use those colors [@elving_2014_npr], it is not clear if this was a deliberate choice or whether this was yet another symptom of an increasingly polarized public who just want easily digestable information about politics so they can move on to other matters in their lives. An undertaking to address such a question is for a different book, perhaps.

To give a preview for this chapter, I propose two studies to explore the claims I make in this chapter. Yard signs are a simple yet effective source of campaign branding for political candidates as they convey information that is often useful for voters -- partisanship [@campbell_et-al_1960_jws]. As yard signs are a simple and static form of campaign advertising, they cannot rely heavily on more complicated forms of information. However, one form of information they rely on and vary in is their color. With these characteristics, yard signs are an ideal candidate both from a statistical standpoint and also substantive for an experimental test of my claims in this chapter. From a statistical standpoint, as they are static forms of party branding they have fewer moving parts, which help isolate the effect that color has on shaping attitudes. From a substantive standpoint, while we understand that yard signs matter to campaigns, we have yet to systematically explore more simple design choices such as color.

In [Study 1](#sec-study-1), I conduct an online experiment that examines whether the public notice color and whether their attitudes are shaped by "Republican Red" and "Democratic Blue" on the yard signs of a fictional candidate. Based on informal conversations with campaign marketing consultants, it appears that campaigns are aware that the colors red and blue betray the partisanship of a candidate to voter. So in [Study 2](#sec-study-2), I examine whether campaigns react to the partisan composition of an electorate as a response to these individual-level patterns. I do this by examining how electoral context shapes variation in using these partisan colors. While these two studies highlight endogenous processes, I am not necessarily arguing that either of these patterns only flow one way. In fact, I believe that there is a necessary feedback loop between the parties and their base of support. Voters support candidates who signal strong allegiance to the party to increase their chances for electoral success [@utych_2020_es] and the parties are motivated to distinguish themselves [@lee_2009;@clifford_2020_pb] as a result of their high levels of ideological and affective polarization [@dietrich_2021_pa;@enders_2021_jop]. Again, to tease out the direction of blame is more appropriate for another time.

@@

A central theme of an edited issue of *The International Journal of Press/Politics* is that visual politics is important and understudied [@lilleker_2019_pm]. Those who are engaged in these questions attribute these challenges to methodological sophistication and the difficult task of interdisciplinary theorizing [@gerodimos_2019_pm; @bucy_joo_2021_ijpp].

The rise of television consumption changed the focus of the medium in the research for political communication scholars [@hall-jamieson_2014_ohpc]. In a similar way, so does the rise of image-based social media and has set new agendas. For several methodological and disciplinary reasons, the visual aspects of television were of little focus in the literature [@bucy_joo_2021_ijpp]. However, the news organizations and politicians are responding to the ubiquitous use of the public to use image-dominant social media platforms like TikTok and Instagram, by joining such platforms. They are relatively active on these platforms as well. Scholars need to make this transition to integrate the role of simple visual information into our theories of political information processing (abbreviated as *pip*) and attitude formation.

How does politically-relevant visual information matter to politics? From an evolutionary-biological perspective, visual information has been a common source of information for millions of years that a variety of single-and-multi-cell organisms rely upon to evaluate their environment [see @grabe_bucy_2009_oup, Chapter 1 for a useful discussion]. Visual information processing, as an ancient biological invention, the human brain is organized around the processing of it. Reflecting on this, many scholars of neuroscience argue that visual information is the fastest form of information processing for humans. For example, even complex visual information, such as the warmth expressed in someone's facial features, is automatically and pre-consciously processed in only about 33 milliseconds (abbreviated as ms) [@ames_et-al_2012_ohsn].

Approached from a different perspective, as humans are cognitive misers \index{cognitive misers}, visual information in the realm of politics provides efficient information to voters about politically-relevant actors and events [@lilleker_2019_pm]. Evidence suggests that even simple party branding on yard signs has an emotional appeal and encourages a shift in attitudes toward the person who owns the yard sign [@makse_et-al_2019_oup]. That is, even branding as simple as yard signs appears to influence the attitudes of those who view them. However, like all types of party branding, there remain questions about the individual visual components' contribution.

More simple forms of visual information are likely even more efficient forms of information. In the context of the United States, the "Republican red" and "Democratic blue" is a relatively recent invention that likely has significant import in an era of significant effort by the parties to distinguish themselves from each other [@clifford_2020_pb] and where voters toe the party line [@utych_2020_es]. Since the 2000 presidential election, the media have consistently used red on their electoral maps in "horserace" journalism to represent Republicans and blue to represent Democrats [@elving_2014_npr]. The supposed consequence is that Democrats now report a preference for the color blue over the color red, and Republicans report a preference for the color red over the color blue [@schloss_palmer_2014_pbr]. Others have demonstrated the strategic choices that candidates make on branding choices -- including color -- based on the types of information they want to convey to voters to distinguish them from their rivals [see @williams_et-al_2022_jomp].

In western Europe, voters are better at connecting the ideological positioning of a party with the color they use in their branding [@casiraghi_et-al_2022_pp]. The ability to do so is strengthened when the party is longer-surviving and is more prominent [@casiraghi_et-al_2022_pp]. The use of politically-relevant colors activates biases toward pre-existing ideological and partisan preferences among voters in a Spanish sample [@maestre_medero_2022_pr]. And male and female candidates in the United States use different colors -- among other things such as fonts -- in their campaign branding to convey distinguishing information about themselves [@williams_et-al_2022_jomp]. What remains unclear is whether these patterns in the United States are just simply patterns or whether there are identifiable causes to these patterns. While political psychologists are increasingly interdisciplinary and rely on psychology and neuroscience, it is relatively unclear how color can convey information as complex as politics. Further, it is unclear whether such a simple type of information has enough efficacy to shape political outcomes such as voting intention. This chapter is a first cut at addressing these open questions.

@@

Color and other simple visual information are processed much quicker and occur more frequently than text-based information that may be communicated through a news article [@mehta_zhu_2009_s]. As the color and other visual information are processed differently, we should also consider its use as political information. As visual information is affectively encoded [@cimbalo_et-al_1978_jgp], it can affect the affective state and processing of more complex information, such as text. The visual information provides a snap judgment or an impression of the object through faster processing and activates particular neurological processes that influence subsequent information appraisals [@ames_et-al_2012_ohsn]. This implies that snap-judgments from visual forms of political information may have downstream effects that shape the way in which we engage with forms of political information that we traditionally consider.

Before expanding upon the role of colors in shaping political attitudes, let me first define an attitude as a concept. An attitude represents an accessible, valanced evaluation of associated prior information and experiences. This conceptualization fits with that of the Object-Evaluation Associations Model [@fazio_2007_sc]. As opposed to viewing attitudes as a latent collection of memories, it views attitudes as measurable evaluations of memories. As memories are at the core of an attitude, the association of memories with its evaluative component [see @kensinger_fields_2022_ohhum] contributes to the perspective that attitudes are affective. This implies that we should be able to measure attitudes but that such an operationalization requires careful consideration of the context's role in any given measure of an attitude as they result from memories [@fazio_2007_sc]. In this chapter, I consider colors' effect on shaping attitudes about a political object -- a yard sign. As I hold this constant in the chapter, concerns about variation in what people are thinking about when seeing the color red and blue is less of a concern in this chapter. Later chapters grapple with this consideration.

In line with existing theories of *pip*, I conceptualize attitudes as associative. This means that attitudes may be unstable - not randomly, in any case. As attitudes are associative, they manifest slightly differently depending on the associative paths activated [@fazio_2007_sc]. The retrieval of relevant memories to the attitude depends on many factors, such as the recency of the event, the similarity of the context, and the importance or salience of the memory [@kahana_et-al_2022_ohhum]. This means that the memories retrieved to contribute to an attitude are quite variable. However, to understand where that variability comes from, we must understand the deeper processes influencing how information is encoded and later retrieved. This illustrates my need to build upon the reigning models of *pip*. Colors may act as a contextual feature that may lead to this variability in how a given set of political information may shape attitudes.

Colors are associative and are affectively encoded [@cimbalo_et-al_1978_jgp]. When individuals access a memory, they do not just recall an object, but they may recall visual information such as the color of an object [@mehta_zhu_2009_s]. Due to these features, colors can be processed pre-consciously and very quickly [@mehta_zhu_2009_s]. As they are affectively encoded, their associations with particular memory contribute to the evaluative component of the memory. For example, individuals associate colors like red with anger and arousal [@valdez_mehrabian_1994], whereas individuals associate blue with things like happiness and pleasure [@dandrade_egan_1974_ae]. This means that colors are compelling as contextual information shaping the subsequent processing and integration of "traditional" forms of political information to construct an attitude. As the preference for a particular color correlates with political attitudes [@schloss_palmer_2014_pbr], they may have some causal influence upon political attitudes; rather than just a correlation with them.

According to the literature in affective neuroscience, visual information is processed in various parts of the brain, such as the visual cortex [@goldstein_brockmole_2017_cl]. These types of processes do not occur in particular areas of the brain as some like to think. As attitudes are associative, components of our brain works as this complex highway connecting different types of information. As a result, the processing of visual information will activate other areas of the brain and will make associated paths "hot." One such area is the amygdala. Neuroscientists believe that as visual information is quickly and subconsciously processed, the amygdala takes and appraises it based on the paths it activated; this generates a simple affective response to such information [@winkielman_et-al_2011_ohsn]. More complex, categorical emotion occurs later in conscious processing [@winkielman_et-al_2011_ohsn].

To provide a less technical description of the processes occuring here, once politically-relevant visual information is detected, the retina passes it to the brain. Once there, the brain attempts to classify the visual information by activating networks of neurons associated with the current information. With these activated pathways, the brain also attempts to appraise such information based on quick classification. For attitude formation, this implies that, as memories are affectively encoded [@kensinger_fields_2022_ohhum], these memories help areas such as the amygdala to appraise the current information.

These fast affective classifications are valanced rather than the more laborious categorization of specific emotions (e.g., anger, anxiety, fear, happy). What this means is that rather than conceptualizing the affective component of this process in line with the popular conceptualizations in political science, such as affective intelligence theory [@marcus_2000_arps], I conceptualize affect as a more extensive system that considers emotion only as a conscious classification process that occurs later than initial valanced appraisals of an object [@sander_2013_chhan; @ralph_anderson_2018_pup]. That is, emotion -- the complex classification of appraisals -- is a conscious component of affect. The pre-conscious process of affect occurs first with the simple and automatic valanced classification of an object [@winkielman_et-al_2011_ohsn; @dror_2017_er].

The pre-conscious appraisals of the visual information one encounters encourages particular behavioral and attitudinal motivations [@valentino_et-al_2011_jop; @ralph_anderson_2018_pup]. This has evolutionary roots for survival [@ralph_anderson_2018_pup; @parker_2003_p]. While affective appraisals can lead to complex motivations, such as anxiety leading to motivations for information seeking [@marcus_2000_arps], affective appraisals are valanced and are more automatic [@winkielman_et-al_2011_ohsn]. These affective appraisals lead to a desire to either retract or engage more with the object [@valentino_et-al_2011_jop]. The snap judgment resulting from the automatic processing of politically relevant visual information should likely lead to an affective response that motivates either a desire to engage more with the object or disengage.

While the visual information may encourage a particular immediate reaction to engage or disengage from the information, subsequent information processing and more conscious processing adjusts this initial appraisal generated by the snap judgment [@kensinger_fields_2022_ohhum]. While subsequent information may amend one's snap-judgment, the snap-judgment nevertheless influences the processing of subsequent information by activating particular paths, which is later encoded as associated with the object as it converts to a memory [@lodge_taber_2013_cup; @kensinger_fields_2022_ohhum].

Beyond a motivated backlash effect, the popular application of motivated reasoning to political science finds evidence suggesting that individuals take more time processing information that runs against their pre-existing preferences [@taber_lodge_2006_ajps]. The explanation for this is that people are finding ways to come up with counterarguments to such information. Therefore, it is possible that we may expect that those interacting with visual information of a political party for which they disagree with spend more time looking at those objects.

@@

Let me illustrate the snap-judgment model with a common experience for residents of the United States. Say you are driving down a highway. At 65 miles an hour, you are traveling at about 95 feet per second at this speed. You split your attention. You focus your eyes on the conditions of the road in front of you, the cars in front of you, and the rear-view mirror where your kids are either dropping food in the crevice between the seats or trying to grab your attention. Out of the corner of your eye, you see a sign. It is not a road sign because it is not on the familiar white or yellow background with black lettering. It is election season. You correctly infer that it is a political yard sign. In this split second, you notice the sign's color and may see a name: Mitch McConnell. You now are racking your brain to think about who that is. If you are politically engaged, you might come to that recognition of the name quickly or it may take you significantly longer if you are less politically engaged [see @kahana_et-al_2022_ohhum]. You figure out that they are a Republican politician. You may have come to this with the help of the fact that every year you have seen yard signs on this stretch of highway. You know that when you see those electoral maps pop up on your news app on your phone that the electoral forecasts always represent Republican support with red and blue for Democrats. Once you have figured out who this person is, with the help of this other information, you react: "ugh, that guy is too loyal to Trump" or "yeah! He's loyal to Trump". You've expressed a political attitude.

What the snap-judgment model predicts is happening in your head is that as soon as the light that bounces off the sign to produce a particular wavelength hits your eyes, your brain is already trying to make sense of this information. This is a valuable tool for survival that biology has optimized for millions of years [@parker_2003_p]. Rather than slowly processing the visual information and finding yourself in the jaws of a predator or processing it quickly but forming the wrong impression and running away from a friend, the brain processes the information quickly and subconsciously [@newell_1990_hup]. To make sense of such information, it accesses familiar information similar to what it is currently attempting to process for efficiency [@kahana_et-al_2022_ohhum]. This is accessing memories and contains valanced information [@kensinger_fields_2022_ohhum]: should I avoid this, or is it pleasant? Once the brain has finished such processing, it can pass its prediction to your conscious memory. Once you form a reflex of avoid or approach, this opens up space for your brain to process the more complex information: to take the patterns of the light as shapes that construct symbols and letters. This comes later because this information not only requires access to information about what it *is* but also what it *means*; once you understand what it means, you have the information necessary to evaluate it.

The snap-judgment model predicts that you first process the colors of the yard sign. You access associative memory to figure out what those particular wavelengths represent: red, white, blue? As these colors are associated with different emotional states [see @cimbalo_et-al_1978_jgp] and the resulting behavioral consequences, your brain starts sending signals to the rest of your body to prepare it to react [see @sander_2013_chhan; @dror_2017_er]. You now need to figure out what the rest of that information was. What were the patterns of that light? It appears that there were some white letters on the sign. There was an "E," a "L," an "E," a "C," and a "T." That creates the word "ELECT." Meaning to vote for. There were some more letters on the sign: an "M," an "I," a "T," a "C," and an "H." A name. The full name is "Mitch McConnell". Since it is about politics, it must be a politician named Mitch McConnell. Now imagine the information was the same, except the color was blue. You may take more time to figure out how that Mitch McConnell person is and come to your reaction to seeing their yard sign. This is because, without the color red, you first are thinking about Democrats who are named Mitch McConnell. Only when you come up empty on your mental Rolodex do you figure out that it is the Republican Mitch McConnell.

How do you react to the color and then to the name? Social groupings are not simply abstract concepts invented by social psychologists; our neurobiology reflects them. For example, researchers find activation of parts of the brain, such as the anterior insula, when we see someone in our social group outperformed by someone from the out-group [see @zink_barter_2012_ohsn]. The anterior insula activity is associated with physical and emotional pain, not just for ourselves but also for others [@adolphs_janowski_2011_ohsn]. Others have also observed that when seeing someone part of a high-status social group, there is an increase in activity in the sensorimotor cortex and supplementary motor area, indicating more activity in the areas of the brain that encourage movement [@zink_et-al_2008_n]. Visual information about someone in your social group speeds up processing, is more salient, and demands more attention than visual information about an object outside your social group [@zink_barter_2012_ohsn].

There is significant evidence supporting the theory that our partisan identification reflects more than just our attitudes about politics but a social identity [see @campbell_et-al_1960_jws; @mason_2018_cup] that guides our attitudes [see @achen_bartels_2016_pup; @white_et-al_2014_apsr; also @bullock_2011_apsr]. As our political attitudes reflect shared views among co-partisans [@pickup_et-al_2020_pb], the congruence between political information and our partisan identification influences our reactions to such political information. This means that the visual information we glean from politics will likely motivate those neurological features of social groups and explain the resulting behavioral manifestations reacting to such information. That information is also likely to be processed at different rates as well. That is, while visual information carries general affective associations, we should also expect associations between politically-relevant colors with partisan identification.

@@

From this discussion, I expect the following: that the average potential voter pays attention to the colors used in campaign branding ($H_{1}$); that these colors that they notice shape perceptions about the person and ideological symbol represented in the branding -- meaning that they express different levels of preference for receiving more information that is similar to what they saw and their levels of preference for supporting such a campaign ($H_{2}$); that the consistency of information explains more positive perceptions -- simple visual information with more complex "traditional" information ($H_{3}$); this positive and consistent information is processed quicker than negative and inconsistent, negative and consistent, and positive and inconsistent information ($H_{4}$); and finally that campaigns make strategic choices about their branding to attract voters ($H_{5}$) in line with their primary objective of reelection [@fenno_1973_lb; @mayhew_1974_yup]. A summary of these hypotheses are included in @tbl-hypotheses.

>>>>

This chapter applies the snap-judgment model to a simple case. The use of color on yard signs used by campaigns in elections. I focus on yard signs as they are a common form of campaign marketing and evidence suggests that they can shape individual behavior [@makse_politics_2019] and electoral outcomes [@green_effects_2016]. Additionally, I start by examining the colors red and blue. While there are other colors that may provide political information -- black and white for Black Lives Matter and Blue Lives Matter movements, the overarching goal of this book is to lay the foundation for a larger research agenda in visual political psychology [see @bucy_editors_2021]. Without such foundations, we are studying complex cases and circumstances upon assumptions without a grasp of how, and if they even do, occur in the cases that might seem trivial to some. While there indeed is empirical and experiential evidence suggesting that the Republicans use the color red and Democrats use the color blue [@schloss_politics_2014;@williams_whats_2022], there is yet to be a theoretical framework that explains how these connections happen at the individual level. It may also be unsatisfying that I do not address *how* these connections came about. While the connections between Republicans with the color red and Democrats with the color blue seemed to quicken after the 2000 Presidential election where news networks covering the election began to consistently use those colors [@elving_color_2014], it is not clear if this was a deliberate choice or whether this was yet another symptom of an increasingly polarized public who just want easily digestible information about politics so they can move on to other matters in their lives. An undertaking to address such a question is for a different book, perhaps.

To give a preview for this chapter, I propose two studies to explore the claims I make in this chapter. Yard signs are a simple yet effective source of campaign branding for political candidates as they convey information that is often useful for voters -- partisanship [@campbell_american_1969]. As yard signs are a simple and static form of campaign advertising, they cannot rely heavily on more complicated forms of information. However, one form of information they rely on and vary in is their color. With these characteristics, yard signs are an ideal candidate both from a statistical standpoint and also substantive for an experimental test of my claims in this chapter. From a statistical standpoint, as they are static forms of party branding they have fewer moving parts, which help isolate the effect that color has on shaping attitudes. From a substantive standpoint, while we understand that yard signs matter to campaigns, we have yet to systematically explore more simple design choices such as color.

In [Study 1](#sec-study-1), I conduct an online experiment that examines whether the public notice color and whether their attitudes are shaped by "Republican Red" and "Democratic Blue" on the yard signs of a fictional candidate. Based on informal conversations with campaign marketing consultants, it appears that campaigns are aware that the colors red and blue betray the partisanship of a candidate to voter. So in [Study 2](#sec-study-2), I examine whether campaigns react to the partisan composition of an electorate as a response to these individual-level patterns. I do this by examining how electoral context shapes variation in using these partisan colors. While these two studies highlight endogenous processes, I am not necessarily arguing that either of these patterns only flow one way. In fact, I believe that there is a necessary feedback loop between the parties and their base of support. Voters support candidates who signal strong allegiance to the party to increase their chances for electoral success [@utych_voter-centric_2020] and the parties are motivated to distinguish themselves [@lee_insecure_2016;@clifford_how_2020] as a result of their high levels of ideological and affective polarization [@dietrich_using_2021;@enders_issues_2021]. Again, to tease out the direction of blame is more appropriate for another time.

@@

A central theme of an edited issue of *The International Journal of Press/Politics* is that visual politics is important and understudied [@lilleker_introduction_2019 ]. Those who are engaged in these questions attribute these challenges to methodological sophistication and the difficult task of interdisciplinary theorizing [@gerodimos_interdisciplinary_2019; @bucy_editors_2021].

The rise of television consumption changed the focus of the medium in the research for political communication scholars [@hall_jamieson_creating_2014]. In a similar way, so does the rise of image-based social media and has set new agendas. For several methodological and disciplinary reasons, the visual aspects of television were of little focus in the literature [@bucy_editors_2021]. However, the news organizations and politicians are responding to the ubiquitous use of the public to use image-dominant social media platforms like TikTok and Instagram, by joining such platforms. They are relatively active on these platforms as well. Scholars need to make this transition to integrate the role of simple visual information into our theories of political information processing (abbreviated as *pip*) and attitude formation.

How does politically-relevant visual information matter to politics? From an evolutionary-biological perspective, visual information has been a common source of information for millions of years that a variety of single-and-multi-cell organisms rely upon to evaluate their environment [see @grabe_image_2009, Chapter 1 for a useful discussion]. Visual information processing, as an ancient biological invention, the human brain is organized around the processing of it. Reflecting on this, many scholars of neuroscience argue that visual information is the fastest form of information processing for humans. For example, even complex visual information, such as the warmth expressed in someone's facial features, is automatically and pre-consciously processed in only about 33 milliseconds (abbreviated as ms) [@ames_impression_2012].

Approached from a different perspective, as humans are cognitive misers \index{cognitive misers}, visual information in the realm of politics provides efficient information to voters about politically-relevant actors and events [@lilleker_power_2019]. Evidence suggests that even simple party branding on yard signs has an emotional appeal and encourages a shift in attitudes toward the person who owns the yard sign [@makse_politics_2019]. That is, even branding as simple as yard signs appears to influence the attitudes of those who view them. However, like all types of party branding, there remain questions about the individual visual components' contribution.

More simple forms of visual information are likely even more efficient forms of information. In the context of the United States, the "Republican red" and "Democratic blue" is a relatively recent invention that likely has significant import in an era of significant effort by the parties to distinguish themselves from each other [@clifford_how_2020] and where voters toe the party line [@utych_voter-centric_2020]. Since the 2000 presidential election, the media have consistently used red on their electoral maps in "horserace" journalism to represent Republicans and blue to represent Democrats [@elving_color_2014]. The supposed consequence is that Democrats now report a preference for the color blue over the color red, and Republicans report a preference for the color red over the color blue [@schloss_politics_2014]. Others have demonstrated the strategic choices that candidates make on branding choices -- including color -- based on the types of information they want to convey to voters to distinguish them from their rivals [see @williams_whats_2022].

In western Europe, voters are better at connecting the ideological positioning of a party with the color they use in their branding [@casiraghi_colors_2022]. The ability to do so is strengthened when the party is longer-surviving and is more prominent [@casiraghi_colors_2022]. The use of politically-relevant colors activates biases toward pre-existing ideological and partisan preferences among voters in a Spanish sample [@losada_maestre_color_2022]. And male and female candidates in the United States use different colors -- among other things such as fonts -- in their campaign branding to convey distinguishing information about themselves [@williams_whats_2022]. What remains unclear is whether these patterns in the United States are just simply patterns or whether there are identifiable causes to these patterns. While political psychologists are increasingly interdisciplinary and rely on psychology and neuroscience, it is relatively unclear how color can convey information as complex as politics. Further, it is unclear whether such a simple type of information has enough efficacy to shape political outcomes such as voting intention. This chapter is a first cut at addressing these open questions.

@@

Color and other simple visual information are processed much quicker and occur more frequently than text-based information that may be communicated through a news article [@mehta_blue_2009]. As the color and other visual information are processed differently, we should also consider its use as political information. As visual information is affectively encoded [@cimbalo_emotionally_1978], it can affect the affective state and processing of more complex information, such as text. The visual information provides a snap judgment or an impression of the object through faster processing and activates particular neurological processes that influence subsequent information appraisals [@ames_impression_2012]. This implies that snap-judgments from visual forms of political information may have downstream effects that shape the way in which we engage with forms of political information that we traditionally consider.

Before expanding upon the role of colors in shaping political attitudes, let me first define an attitude as a concept. An attitude represents an accessible, valanced evaluation of associated prior information and experiences. This conceptualization fits with that of the Object-Evaluation Associations Model [@russell_h_fazio_attitudes_2007]. As opposed to viewing attitudes as a latent collection of memories, it views attitudes as measurable evaluations of memories. As memories are at the core of an attitude, the association of memories with its evaluative component [see @kensinger_affective_2022] contributes to the perspective that attitudes are affective. This implies that we should be able to measure attitudes but that such an operationalization requires careful consideration of the context's role in any given measure of an attitude as they result from memories [@russell_h_fazio_attitudes_2007]. In this chapter, I consider colors' effect on shaping attitudes about a political object -- a yard sign. As I hold this constant in the chapter, concerns about variation in what people are thinking about when seeing the color red and blue is less of a concern in this chapter. Later chapters grapple with this consideration.

In line with existing theories of *pip*, I conceptualize attitudes as associative. This means that attitudes may be unstable - not randomly, in any case. As attitudes are associative, they manifest slightly differently depending on the associative paths activated [@russell_h_fazio_attitudes_2007]. The retrieval of relevant memories to the attitude depends on many factors, such as the recency of the event, the similarity of the context, and the importance or salience of the memory [@kahana_laws_2022]. This means that the memories retrieved to contribute to an attitude are quite variable. However, to understand where that variability comes from, we must understand the deeper processes influencing how information is encoded and later retrieved. This illustrates my need to build upon the reigning models of *pip*. Colors may act as a contextual feature that may lead to this variability in how a given set of political information may shape attitudes.

Colors are associative and are affectively encoded [@cimbalo_emotionally_1978]. When individuals access a memory, they do not just recall an object, but they may recall visual information such as the color of an object [@mehta_blue_2009]. Due to these features, colors can be processed pre-consciously and very quickly [@mehta_blue_2009]. As they are affectively encoded, their associations with particular memory contribute to the evaluative component of the memory. For example, individuals associate colors like red with anger and arousal [@valdez_effects_1994], whereas individuals associate blue with things like happiness and pleasure [@dandrade_colors_1974]. This means that colors are compelling as contextual information shaping the subsequent processing and integration of "traditional" forms of political information to construct an attitude. As the preference for a particular color correlates with political attitudes [@schloss_politics_2014], they may have some causal influence upon political attitudes; rather than just a correlation with them.

According to the literature in affective neuroscience, visual information is processed in various parts of the brain, such as the visual cortex [@goldstein_sensation_2017]. These types of processes do not occur in particular areas of the brain as some like to think. As attitudes are associative, components of our brain works as this complex highway connecting different types of information. As a result, the processing of visual information will activate other areas of the brain and will make associated paths "hot." One such area is the amygdala. Neuroscientists believe that as visual information is quickly and subconsciously processed, the amygdala takes and appraises it based on the paths it activated; this generates a simple affective response to such information [@winkielman_emotion_2011]. More complex, categorical emotion occurs later in conscious processing [@winkielman_emotion_2011].

To provide a less technical description of the processes occuring here, once politically-relevant visual information is detected, the retina passes it to the brain. Once there, the brain attempts to classify the visual information by activating networks of neurons associated with the current information. With these activated pathways, the brain also attempts to appraise such information based on quick classification. For attitude formation, this implies that, as memories are affectively encoded [@kensinger_affective_2022], these memories help areas such as the amygdala to appraise the current information.

These fast affective classifications are valanced rather than the more laborious categorization of specific emotions (e.g., anger, anxiety, fear, happy). What this means is that rather than conceptualizing the affective component of this process in line with the popular conceptualizations in political science, such as affective intelligence theory [@marcus_emotions_2000], I conceptualize affect as a more extensive system that considers emotion only as a conscious classification process that occurs later than initial valanced appraisals of an object [@sander_cambridge_2013; @ralph_neuroscience_2018]. That is, emotion -- the complex classification of appraisals -- is a conscious component of affect. The pre-conscious process of affect occurs first with the simple and automatic valanced classification of an object [@winkielman_emotion_2011; @dror_deconstructing_2017].

The pre-conscious appraisals of the visual information one encounters encourages particular behavioral and attitudinal motivations [@valentino_election_2011; @ralph_neuroscience_2018]. This has evolutionary roots for survival [@ralph_neuroscience_2018; @parker_blink_2003]. While affective appraisals can lead to complex motivations, such as anxiety leading to motivations for information seeking [@marcus_emotions_2000], affective appraisals are valanced and are more automatic [@winkielman_emotion_2011]. These affective appraisals lead to a desire to either retract or engage more with the object [@valentino_election_2011]. The snap judgment resulting from the automatic processing of politically relevant visual information should likely lead to an affective response that motivates either a desire to engage more with the object or disengage.

While the visual information may encourage a particular immediate reaction to engage or disengage from the information, subsequent information processing and more conscious processing adjusts this initial appraisal generated by the snap judgment [@kensinger_affective_2022]. While subsequent information may amend one's snap-judgment, the snap-judgment nevertheless influences the processing of subsequent information by activating particular paths, which is later encoded as associated with the object as it converts to a memory [@lodge_rationalizing_2013; @kensinger_affective_2022].

Beyond a motivated backlash effect, the popular application of motivated reasoning to political science finds evidence suggesting that individuals take more time processing information that runs against their pre-existing preferences [@taber_motivated_2006]. The explanation for this is that people are finding ways to come up with counterarguments to such information. Therefore, it is possible that we may expect that those interacting with visual information of a political party for which they disagree with spend more time looking at those objects.

@@

Let me illustrate the snap-judgment model with a common experience for residents of the United States. Say you are driving down a highway. At 65 miles an hour, you are traveling at about 95 feet per second at this speed. You split your attention. You focus your eyes on the conditions of the road in front of you, the cars in front of you, and the rear-view mirror where your kids are either dropping food in the crevice between the seats or trying to grab your attention. Out of the corner of your eye, you see a sign. It is not a road sign because it is not on the familiar white or yellow background with black lettering. It is election season. You correctly infer that it is a political yard sign. In this split second, you notice the sign's color and may see a name: Mitch McConnell. You now are racking your brain to think about who that is. If you are politically engaged, you might come to that recognition of the name quickly or it may take you significantly longer if you are less politically engaged [see @kahana_laws_2022]. You figure out that they are a Republican politician. You may have come to this with the help of the fact that every year you have seen yard signs on this stretch of highway. You know that when you see those electoral maps pop up on your news app on your phone that the electoral forecasts always represent Republican support with red and blue for Democrats. Once you have figured out who this person is, with the help of this other information, you react: "ugh, that guy is too loyal to Trump" or "yeah! He's loyal to Trump". You've expressed a political attitude.

What the snap-judgment model predicts is happening in your head is that as soon as the light that bounces off the sign to produce a particular wavelength hits your eyes, your brain is already trying to make sense of this information. This is a valuable tool for survival that biology has optimized for millions of years [@parker_blink_2003]. Rather than slowly processing the visual information and finding yourself in the jaws of a predator or processing it quickly but forming the wrong impression and running away from a friend, the brain processes the information quickly and subconsciously [@newell_unified_1990]. To make sense of such information, it accesses familiar information similar to what it is currently attempting to process for efficiency [@kahana_laws_2022]. This is accessing memories and contains valanced information [@kensinger_affective_2022]: should I avoid this, or is it pleasant? Once the brain has finished such processing, it can pass its prediction to your conscious memory. Once you form a reflex of avoid or approach, this opens up space for your brain to process the more complex information: to take the patterns of the light as shapes that construct symbols and letters. This comes later because this information not only requires access to information about what it *is* but also what it *means*; once you understand what it means, you have the information necessary to evaluate it.

The snap-judgment model predicts that you first process the colors of the yard sign. You access associative memory to figure out what those particular wavelengths represent: red, white, blue? As these colors are associated with different emotional states [see @cimbalo_emotionally_1978] and the resulting behavioral consequences, your brain starts sending signals to the rest of your body to prepare it to react [see @sander_cambridge_2013; @dror_deconstructing_2017]. You now need to figure out what the rest of that information was. What were the patterns of that light? It appears that there were some white letters on the sign. There was an "E," a "L," an "E," a "C," and a "T." That creates the word "ELECT." Meaning to vote for. There were some more letters on the sign: an "M," an "I," a "T," a "C," and an "H." A name. The full name is "Mitch McConnell". Since it is about politics, it must be a politician named Mitch McConnell. Now imagine the information was the same, except the color was blue. You may take more time to figure out how that Mitch McConnell person is and come to your reaction to seeing their yard sign. This is because, without the color red, you first are thinking about Democrats who are named Mitch McConnell. Only when you come up empty on your mental Rolodex do you figure out that it is the Republican Mitch McConnell.

How do you react to the color and then to the name? Social groupings are not simply abstract concepts invented by social psychologists; our neurobiology reflects them. For example, researchers find activation of parts of the brain, such as the anterior insula, when we see someone in our social group outperformed by someone from the out-group [see @zink_neural_2012]. The anterior insula activity is associated with physical and emotional pain, not just for ourselves but also for others [@adolphs_emotion_2011]. Others have also observed that when seeing someone part of a high-status social group, there is an increase in activity in the sensorimotor cortex and supplementary motor area, indicating more activity in the areas of the brain that encourage movement [@zink_know_2008]. Visual information about someone in your social group speeds up processing, is more salient, and demands more attention than visual information about an object outside your social group [@zink_neural_2012].

There is significant evidence supporting the theory that our partisan identification reflects more than just our attitudes about politics but a social identity [see @campbell_american_1969; @mason_uncivil_2018] that guides our attitudes [see @achen_democracy_2016; @white_selling_2014; also @bullock_elite_2011]. As our political attitudes reflect shared views among co-partisans [@pickup_expressive_2020], the congruence between political information and our partisan identification influences our reactions to such political information. This means that the visual information we glean from politics will likely motivate those neurological features of social groups and explain the resulting behavioral manifestations reacting to such information. That information is also likely to be processed at different rates as well. That is, while visual information carries general affective associations, we should also expect associations between politically-relevant colors with partisan identification.

@@

From this discussion, I expect the following: that the average potential voter pays attention to the colors used in campaign branding ($H_{1}$); that these colors that they notice shape perceptions about the person and ideological symbol represented in the branding -- meaning that they express different levels of preference for receiving more information that is similar to what they saw and their levels of preference for supporting such a campaign ($H_{2}$); that the consistency of information explains more positive perceptions -- simple visual information with more complex "traditional" information ($H_{3}$); this positive and consistent information is processed quicker than negative and inconsistent, negative and consistent, and positive and inconsistent information ($H_{4}$); and finally that campaigns make strategic choices about their branding to attract voters ($H_{5}$) in line with their primary objective of reelection [@fenno_congressmen_1973; @mayhew_congress_1974]. A summary of these hypotheses are included in @tbl-hypotheses.
